---
title: "Exercise 4"
subtitle: "Bootstraping Practicals I"
author: "Olesia Galynskaia 12321492"
date: "2025"
output: pdf_document
---

# Fix the random seed for reproducibility
```{r seed, echo=TRUE}
set.seed(12321492)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 6, fig.height = 4, dpi = 150)
```

# Task 7

Making a sample
```{r}
x_t7 <- c(82, 107, 93)
```

## 1. Mean of the sample
The sample contains the values 82, 107, and 93.  
The mean of this sample is:

\[
\bar{x} = \frac{82 + 107 + 93}{3} = 94.
\]

```{r}
mean_t7 <- mean(x_t7)
mean_t7
```

## 2. Number of possible bootstrap samples + list all triplets

We take a bootstrap sample of size 3 with replacement.  
Each position can be 82, 93, or 107.  
This gives \(3^3 = 27\) possible combinations.  
`expand.grid()` prints all 27 triplets.
```{r}
# Generate all possible bootstrap samples (all triplets with replacement)
all_triplets <- expand.grid(x_t7, x_t7, x_t7)

# Count how many there are
n_triplets <- nrow(all_triplets)

n_triplets
all_triplets
```

## 3. Mean of each bootstrap sample
Here we take every bootstrap triplet and compute its mean.  
`rowMeans()` gives the average of each row in the table, so we get 27 mean values — one for each possible bootstrap sample.
```{r}
bootstrap_means <- rowMeans(all_triplets)

bootstrap_means
```

## 4. Mean of all bootstrap means
The average of all 27 bootstrap means is basically the same as the original sample mean.  
Here both values are 94.  
For the sample mean, the bootstrap does not introduce bias, so the two match.

```{r}
orig_mean <- mean(x_t7)
mean_bootstrap_means <- mean(bootstrap_means)

orig_mean
mean_bootstrap_means
```

## 5. Fit a normal distribution to the sample
Here we pretend the data comes from a normal distribution.  
To "fit" this normal distribution, we just take the sample mean and the sample standard deviation.  
These two numbers fully define the normal model.  
```{r}
norm_mean <- mean(x_t7)
norm_sd   <- sd(x_t7)

norm_mean
norm_sd
```

## 6. Draw 27 normal samples of size 3 and compute their means
We use the normal model fitted in step 5 (same mean and sd as the sample).  
Then we generate 27 new samples of size 3 from this normal distribution.  
For each sample we compute the mean.  
These normal-based means should be close to the original mean (94) on average, just like the bootstrap means.
```{r}
n_samples <- 27

normal_means <- replicate(n_samples, mean(rnorm(3, mean = norm_mean, sd = norm_sd)))

normal_means
```

## 7. Low and high values of the means
The bootstrap means go from 82 to 107, because these are the smallest and largest possible values we can get when resampling the original numbers.  
The normal-based means have a wider range (in my run about 74 to 108), because the normal distribution is continuous and can produce values outside the original sample.
```{r}
# Range of bootstrap means (from all 27 triplets)
range_bootstrap <- range(bootstrap_means)

# Range of normal-based means (27 samples generated in step 6)
range_normal <- range(normal_means)

range_bootstrap
range_normal
```

# Task 8

Making a sample
```{r}
x_t8 <- c(4.94, 5.06, 4.53, 5.07, 4.99, 5.16,
          4.38, 4.43, 4.93, 4.72, 4.92, 4.96)

n_t8 <- length(x_t8)
n_t8
```

## 1. Number of possible bootstrap samples
A bootstrap sample has size 12 and we sample with replacement from 12 values.  
Each of the 12 positions can be any of the 12 data points, so the total number of possible bootstrap samples is \(12^{12}\).  

```{r}
n_boot_t8 <- n_t8^n_t8
n_boot_t8
```

## 2. Mean of each bootstrap sample
Here we just compute the usual sample mean of the 12 observations.  
This value will be our reference when we look at the bootstrap means.
```{r}
mean_t8 <- mean(x_t8)
mean_t8
```

## 3. 2000 bootstrap samples and their means
We draw 2000 bootstrap samples from `x_t8`.  
Each bootstrap sample has size 12 and is drawn with replacement.  
For every resample we compute the mean, so `boot_means_t8` contains 2000 bootstrap means.
```{r}
B_t8 <- 2000

# 2000 bootstrap means: each time we resample 12 values with replacement
boot_means_t8 <- replicate(B_t8, mean(sample(x_t8, replace = TRUE)))

# Look at the first few bootstrap means
head(boot_means_t8)
```

We do not create new bootstrap samples here.  
We just average:  
- the first 20 bootstrap means,  
- the first 200 bootstrap means,  
- and all 2000 bootstrap means.  

As we use more bootstrap means, these averages get closer to the sample mean from step 2.

### 3.1 Mean of first 20
```{r}
mean_20   <- mean(boot_means_t8[1:20])

mean_20
```

### 3.2 Mean of first 200
```{r}
mean_200   <- mean(boot_means_t8[1:200])

mean_200
```

### 3.3 Mean of 2000
```{r}
mean_2000 <- mean(boot_means_t8)

mean_2000
```


## 4. Relate all the different bootstrap means to the sample mean
Here we put everything side by side:  
the original sample mean and the three averages of bootstrap means.  

The averages based on 20 and 200 bootstrap means are close to the sample mean but still a bit noisy.  
The mean based on all 2000 bootstrap means is almost identical to the sample mean,
which shows that the bootstrap distribution is centered around it.

```{r}
c(sample_mean = mean_t8,
  mean_20     = mean_20,
  mean_200    = mean_200,
  mean_2000   = mean_2000)
```

## 5. Bootstrap 95% intervals from 20, 200 and 2000 means

```{r}
ci_20   <- quantile(boot_means_t8[1:20],   probs = c(0.025, 0.975))
ci_200  <- quantile(boot_means_t8[1:200],  probs = c(0.025, 0.975))
ci_2000 <- quantile(boot_means_t8,         probs = c(0.025, 0.975))

ci_20
ci_200
ci_2000
```

We use the empirical 2.5% and 97.5% quantiles of the bootstrap means to build 95% intervals.  

With only 20 bootstrap means the interval is quite rough and unstable.  
With 200 bootstrap means the interval is more stable.  
With 2000 bootstrap means the interval changes only slightly compared to 200; it looks much more reliable.  

The intervals from 200 and 2000 means are very similar (roughly [4.70, 4.97]), which shows that 2000 bootstrap samples are enough for a stable estimate.

So more bootstrap samples give a smoother and more trustworthy interval.

## 6. t-test based 95% CI and comparison

```{r}
t_ci_t8 <- t.test(x_t8, conf.level = 0.95)$conf.int
t_ci_t8
```

`t.test()` gives the usual 95% confidence interval for the mean based on a normal / t-distribution assumption.  
In this example t-based interval [4.67, 5.01] is slightly wider than the bootstrap intervals (about [4.71, 4.97] and [4.70, 4.97]).  

The 95% bootstrap intervals from 200 and 2000 means are a bit shorter and sit in a similar region.  
All intervals contain the sample mean and are close to each other,  
so the bootstrap and the t-based method tell a consistent story about the mean.